
%\renewcommand{\thechapter}{\hspace{-0.45cm}}
\addcontentsline{toc}{chapter}{Acknowledgements}
\chapter*{Acknowledgements}
\begin{minipage}[t][0pt]{\linewidth}

Many individuals and organizations enabled and supported me in writing this thesis.\\

 First of all, I thank Prof. J\"{u}rgen Kurths for the continuous trust and support during the last four years and for hosting me at the Potsdam Institute for Climate Impact Research (PIK). Special thanks go to Jobst Heitzig for day-to-day supervision, for numerous discussions, for plenty of freedom when I needed it and for guidance and holding me accountable when I wanted it.
 I want to thank Ricarda Winkelmann and Jean-Denis Mathias for their willingness to critically evaluate this thesis.
 Many thanks to my colleagues and Friends especially Jonathan Donges, Reik Donner, Marc Wiedermann, Wolfram Barfuss and Benjamin Maier, I much appreciated your extensive comments and feedback on my work in various stages. I want to thank them and all of my other colleagues at the COPAN flagship project also for their input and shared experiences in scientific life and work over the last four years.
 Also, I am deeply indebted to my coauthors, especially Finn M\"{u}ller-Hansen, and Doyne Farmer from whom I have learned a great deal about the strengths and weaknesses of economic models, Yuki Asano, who is one of the most industrious persons that I have met in my life and who constantly pushed us to do better, faster and Maurits Ertsen with whom, during long calls on the phone, I had the most interesting discussions on the meaning of modelling and its limits as an ontological tool.\\


 I am much obliged to the Foundation of German Industries (SDW) for placing their trust in me in the early phase of this project. Their scholarship was so much more than just financial support. I also want to thank the Princeton-Humboldt Cooperation and Collective Cognition Network (CoCCoN), and there within especially Pawel Romanczuk, for the opportunity to participate and providing the funds to repeatedly visit Princeton University.
 I am also grateful for all of those who provided me with library and technical infrastructure especially the European Regional Development Fund (ERDF), the German Federal Ministry of Education and Research and the Land Brandenburg for enabling my work by providing resources on the high performance computer systems at PIK.\\

 I am vary grateful to Winnie Poel for many things but at this point especially for her continued support and the big parts of our shared reproductive labor that she took on during the final stages of writing this thesis.
And last but not least, I feel blessed for my two wonderful daughters who came into my life during the work on this thesis and who never get tired of reminding me that there is plenty of life outside of research.
\end{minipage}
